Kindly Bent to Free Us

### 2019-07-07 kindly/lang.tex
Small fixes.
cosmetics-syntax

We now present a formalization of the \lang language featuring
a minimal ML language with let-polymorphism, abstract types, pairs, and our novel
linear type system with borrows which uses kinds and qualified types.
For ease of presentation,
our formalization makes the following simplifications.
First, we only consider matching on pairs, instead of arbitrary algebraic
datatypes. We consider separate operators for borrowing and reborrowing (taking
the borrow of a borrow). We consider regions to be fully annotated
and present a set of rules to infer their placement.
Finally we emulates region variables
as a series of nesting indices.

In the rest of this section, we present the syntax (\cref{syntax})
the syntax directed typing rules (\cref{sdtyping})
and a linearity-aware semantics (\cref{sem}) for \lang.

@@

@@

We illustrate this on the expression $\region{x}{f(\borrow[\IBORROW]{c})}$.
The first step is to check that
$\svar[\IBORROW]{c}{\text{channel}}$ is in $\E$. We then define $\E'$ which
corresponds to $\E$ where the suspended binding is replaced by a borrow binding.
%
This is done by the borrow relation, noted $\lregion{C}{x}{\E}{\E'}$ and
whose main rules are shown in \cref{selectrules:region}.
In our case, we consider an immutable binding, so we replace the suspended
binding by $\bvar{\borrow[\IBORROW]{c}}{\borrowty[\IBORROW]{k}{\text{channel}}}$.
We must also check that $\entail{C}{(\kun_n\lk k\lk\kun_\infty)}$.
We can then typecheck the body of the region, $f(\borrow[\IBORROW]{c})$ in our example, and check that the borrow made available in the region does not escape.
This is done by the constraint $\Cleq{\tau}{\klin_{n-1}}$, which ensures
that the kind of $\tau$ does not contain anything whose kind is ``above $n$''
(for instance, $\borrowty[\IBORROW]{\kun_n}{\text{channel}}$ would have kind
$\kun_n$).

\begin{figure}[!h]
  \begin{minipage}{0.35\linewidth}
    \centering
    \begin{mathpar}
      \ruleSDRegion
    \end{mathpar}
    \caption{The {\sc Region} rule}
    \label{selectrules:region}
  \end{minipage}\hfill
  \begin{minipage}{0.6\linewidth}
    \centering
    \begin{tabular}
      {@{}>{$}r<{$}@{ $\vdash_e$ }
      >{$}c<{$}@{ $\rightsquigarrow_n^{x}$ }
      >{$}l<{$}
      r}

>>>>

\lang is a minimal ML-like language with
let-polymorphism, abstract types, and pairs. Its type system manages
linearity and borrows using kinds and (kind-) constrained types.
For ease of presentation, we keep the language as simple as possible.
\begin{itemize}
\item We showcase pattern matching on pairs rather than algebraic
  datatypes.
\item We have separate operators for borrowing and reborrowing (taking
  the borrow of a borrow).
\item We assume that regions are explicit in the code. (But we present
  an algorithm to infer them in \cref{regionannot}.)
\item We use de Bruijn-indexing to identify regions instead of region
  variables. 
\end{itemize}

The rest of this section formalizes  \lang: the syntax (\cref{syntax}),
the statics in terms of syntax directed typing (\cref{sdtyping}),
and a dynamics that is linearity-aware (\cref{sem}).

@@

\Cref{grammar} defines the syntax of \lang. Expressions are as usual
in an ML-like language.  The novel aspects are match
specifications, regions, and borrows.

We start from the bottom.
A borrow $\borrow{x}$ is always taken from a variable $x$. The
\emph{borrow annotation} $\BORROW$ indicates whether the borrow is mutable
($\MBORROW$) or immutable ($\IBORROW$). If $x$ is already borrowed,
then we need to use the reborrow expression $\reborrow{x}$.

A region $\region{x}{e}$ is annotated with its nesting $n$
and the variable $x$ that may be borrowed in $e$.

A match is indexed with a \emph{match specification} $\etransfm$ that indicates
whether the match should work on borrows ($\etransfm=\&^b$) or not ($\etransfm=id$).

Many types are indexed with kinds, ranged over by $k$.
A kind is either a kind variable $\kvar$ or a constant
(linear $\klin$, affine $\kaff$, or unrestricted $\kun$) indexed
by a nesting level $n \in \Nat \cup \{\infty\}$.

A type $\tau$ is either a type variable, a function type indexed by a
kind, 
a type application $\tapp{t}{\Multi{\tau}}$ of an abstract type constructor
$\T t$, or a borrowed type  $\borrowty{k}{\tau}$.

Type schemes $\schm$ add quantification over kind
variables $\kappa$, kinded type variables $\alpha:k$, and constraints
$C$ to a type, where a constraint is a list of inequalities over
kinds.

Abstract type constructors possess kind schemes $\kschm$ which relate
the kinds of the type constructors arguments to the kind of the
constructed type.

Generally, we write lists with an overbar and (sometimes) an index as
in $\Multi[i]\tau$.

\paragraph{Previously}

@@

We illustrate this on the expression $\region{x}{f(\borrow[\IBORROW]{c})}$
when $c$ is a linear channel.
The first step is to check that
$\svar[\IBORROW]{c}{\text{channel}}$ is in $\E$. We then define $\E'$ which
corresponds to $\E$ where the suspended binding is replaced by a borrow binding.
%
This is done by the borrow relation, noted $\lregion{C}{x}{\E}{\E'}$ and
whose main rules are shown in \cref{selectrules:region}.
In our case, we consider an immutable binding, so we replace the suspended
binding by $\bvar{\borrow[\IBORROW]{c}}{\borrowty[\IBORROW]{k}{\text{channel}}}$.
We must also check that $\entail{C}{(\kun_n\lk k\lk\kun_\infty)}$.
We can then typecheck the body of the region, $f(\borrow[\IBORROW]{c})$ in our example, and check that the borrow made available in the region does not escape.
This is done by the constraint $\Cleq{\tau}{\klin_{n-1}}$, which ensures
that the kind of $\tau$ does not contain anything whose kind is ``above $n$''
(for instance, $\borrowty[\IBORROW]{\kun_n}{\text{channel}}$ would have kind
$\kun_n$).

\begin{figure}[!h]
  \begin{minipage}{0.4\linewidth}
    \centering
    \begin{mathpar}
      \ruleSDRegion
    \end{mathpar}
    \caption{The {\sc Region} rule}
    \label{selectrules:region}
  \end{minipage}\hfill
  \begin{minipage}{0.6\linewidth}
    \centering
    \begin{tabular}
      {@{}>{$}r<{$}@{ $\vdash_e$ }
      >{$}c<{$}@{ $\rightsquigarrow_n^{x}$ }
      >{$}l<{$}
      r}
